---
title: "Class 11: Alpha Fold"
author: "Tonhi Dinh (A18508020)"
format: pdf
toc: TRUE
---

## Background

We saw last day that the main repository of biomolecular structure (the PDB database) onoly has ~250,000 entries.

UnitProtKB (the main protein sequence database) has over 200 Million entries

In this hands-on session we will utilize AlphaFold to predict protein structure from sequence (Jumper et al. 2021).

Without the aid of such approaches, it can take years of expensive laboratory work to determine the structure of just one protein. With AlphaFold we can now accurately compute a typical protein structure in as little as ten minutes.

## The EBI Alphafold Database

The EBI alphafold database contains alots of computed structure models. It is increasingly likely that the structure you are interested in is already in this database at <https://alphafold.ebi.ac.uk>

There are three major outputs from Alphafold

1. A model of structure in **PDB** format
2. A **plDDt score**: that tells us how confident the model is for a given residue in your protein (high value is good, above 70)
3. A **PAE score** that tells us about protein packing quality

If you can't find a matching entry for the sequence you are interested in AFDB you can run Alphafold yourself...

## Running Alphafold

We will use Colabfold to run Alphafold on our sequence <https://github.com/sokrypton/ColabFold?tab=readme-ov-file>

Figure from Alphafold here!

## Interpreting Results

Custom analysis of resulting models

We can read all the Alphafold results into R and do more quantitative analysis than just viewing the structures in Mol-star:

Read all the PDB models
```{r}
library(bio3d)

```

```{r}
pdb_files <- list.files("hivpr_23119/", pattern = ".pdb", full.names = TRUE)
pdb_files
```


```{r}
library(bio3dview)
pdbs <- pdbaln(pdb_files, fit = TRUE, exefile = "msa")
view.pdbs(pdbs)
```

```{r}
# Quick view of model sequence
pdbs
```


```{r}
rd <- rmsd(pdbs, fit=T)
```

```{r}
range(rd)
```


```{r}
library(pheatmap)

colnames(rd) <- paste0("m",1:5)
rownames(rd) <- paste0("m",1:5)
pheatmap(rd)

```


```{r}
# Read a reference PDB structure
pdb <- read.pdb("1hsg")
```
Obtain secondary structure from a call to stride() or dssp() on any of the model structures.

```{r}
plotb3(pdbs$b[1,], typ="l", lwd=2, sse=pdb)
points(pdbs$b[2,], typ="l", col="pink3")
points(pdbs$b[3,], typ="l", col="steelblue")
points(pdbs$b[4,], typ="l", col="darkseagreen")
points(pdbs$b[5,], typ="l", col="salmon2")
abline(v=100, col="gray")
```

We can improve the superposition/fitting of our models by finding the most consistent “rigid core” common across all the models. For this we will use the core.find() function:

```{r}
core <- core.find(pdbs)
```


We can now use the identified core atom positions as a basis for a more suitable superposition and write out the fitted structures to a directory called corefit_structures:

```{r}
core.inds <- print(core, vol=0.5)
```

```{r}
xyz <- pdbfit(pdbs, core.inds, outpath="corefit_structures")
```

Now we can examine the RMSF between positions of the structure. RMSF is an often used measure of conformational variance along the structure:

```{r}
rf <- rmsf(xyz)

plotb3(rf, sse=pdb)
abline(v=100, col="cadetblue", ylab="RMSF")
```

## Predicted Alignment Error for Domains

```{r}
library(jsonlite)

# Listing of all PAE JSON files
pae_files <- list.files(path="hivpr_23119/",
                        pattern=".*model.*\\.json",
                        full.names = TRUE)
```

```{r}
pae1 <- read_json(pae_files[1],simplifyVector = TRUE)
pae5 <- read_json(pae_files[5],simplifyVector = TRUE)

attributes(pae1)
```

```{r}
# Per-residue pLDDT scores 
#  same as B-factor of PDB..
head(pae1$plddt) 
```

```{r}
pae1$max_pae
```

```{r}
pae5$max_pae
```

We can plot the N by N (where N is the number of residues) PAE scores with ggplot or with functions from the Bio3D package:
```{r}
plot.dmat(pae1$pae, 
          xlab="Residue Position (i)",
          ylab="Residue Position (j)")
```

```{r}
plot.dmat(pae5$pae, 
          xlab="Residue Position (i)",
          ylab="Residue Position (j)",
          grid.col = "black",
          zlim=c(0,30))
```

```{r}
#We should really plot all of these using the same z range. Here is the model 1 plot again but this time using the same data range as the plot for model 5:

plot.dmat(pae1$pae, 
          xlab="Residue Position (i)",
          ylab="Residue Position (j)",
          grid.col = "black",
          zlim=c(0,30))
```























