---
title: "Class 5 - Data Viz with ggplot"
author: "Tonhi (PID: 18508020)"
format: pdf
toc: true
---

## Background

There are lots of ways to make figures in R. These include so-called "Base R" graphics (e.g. `plot()`) and tines of add-on packages like **ggplot2**

For example here we make the same plot with both:

```{r}
head(cars)
```

```{r}
plot(cars)
```
First I need to install teh package with the command 
`install.packages()`

> **N.B.** we never run an install cmd in a quarto code chunk or we will end up re-installing packages many many times - this is not what we want!

Every time we want to use one of these "add-on" packages we need to load it up in R with the `library()` function

```{r}
library(ggplot2)
```

```{r}
ggplot(cars)
```

Every ggplot needs at least 3 things:

- The **data**, the stuff you want  plotted
- The **aes**thetics, how the data map to the plot
- The **geom**etry, the type of plot

```{r}
ggplot(cars) + 
  aes(x=speed, y=dist) + 
  geom_point()
```

Add a line to better show relationship between two things
```{r}
p <- ggplot(cars) +
  aes(x=speed, y=dist) +
  geom_point() +
  geom_smooth(method="lm", se=FALSE) + 
  labs(title="Stopping distance of old cars",
       subtitles="Data from cars object",
       x="Speed (mph)", y="Distance (ft)")
```

render it out
```{r}
p + theme_bw()
```




## Gene expression plot

We can read the input data from the class website

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

A first version plot
```{r}
ggplot(genes) + 
  aes(Condition1, Condition2) +
  geom_point()
```

```{r}
table(genes$State)
```

Version 2 let's color by `State` and we can see up and down significant genes compared to all the unchanging genes
```{r}
ggplot(genes) + 
  aes(Condition1, Condition2, col=State) +
  geom_point()
```

Version 3 plot, let's modify the default colors to something we like
```{r}
ggplot(genes) + 
  aes(Condition1, Condition2, col=State) +
  geom_point() + 
  scale_colour_manual(values=c("navy","gray","pink"))+
  labs(x="Control(no drug)", y="Drug", title="Gene Expression Changes upon GLP-1") + 
  theme_bw()

```


## Going further
```{r}
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
```

```{r}
head(gapminder, 3)
```

```{r}
ggplot(data=gapminder) +
  aes(x=gdpPercap, y=lifeExp, col=continent) +
  geom_point(alpha=0.3)
```

Let's "facet" (i.e. make a separate plot) by continent rather than the big hot mess above.
```{r}
ggplot(data=gapminder) +
  aes(x=gdpPercap, y=lifeExp, col=continent) +
  geom_point(alpha=0.3) + 
  facet_wrap(~continent)
```

## Custom plots

How big is this gapminder dataset?

```{r}
nrow(gapminder)
ncol(gapminder)
```
I want to filter down a subset of this fata. I will use **dplyr** package to help me.

First I need to install it and load it up
`install.packages("dplyr")` and then `library("dplyr")`

```{r}
library(dplyr)

gapminder_2007 <- filter(gapminder, year==2007)
head(gapminder_2007)
```

```{r}
filter(gapminder_2007, country=="Ireland")
```

```{r}
filter(gapminder, year==2007, country=="Ireland")
```

```{r}
filter(gapminder, year==2007, country=="United States")
```

> Q. Make a plot comparing 1977 and 2007 for all countries

```{r}
input <- filter(gapminder, year %in% c(1977,2007))
ggplot(data=input) +
  aes(x=gdpPercap, y=lifeExp, col=continent, size=pop) +
  geom_point(alpha=0.3) + 
  facet_wrap(~year)
```













