---
title: "Class 18: Pertussis Mini-Project"
author: Tonhi Dinh (A18508020)
format: pdf
toc: true
---

## Background
Pertusis (whooping cough) is a common lung infection  caused by the bacteria B. Pertussis. It can infect everyone but is most deadly for infacts (under 1 year of age)

## CDC tracking data
The CDC tracks the number of Pertussis cases: 

```{r, echo=FALSE}
cdc <- data.frame(year = c(1922L,1923L,1924L,1925L,
                           1926L,1927L,1928L,1929L,1930L,1931L,
                           1932L,1933L,1934L,1935L,1936L,
                           1937L,1938L,1939L,1940L,1941L,1942L,
                           1943L,1944L,1945L,1946L,1947L,
                           1948L,1949L,1950L,1951L,1952L,
                           1953L,1954L,1955L,1956L,1957L,1958L,
                                          1959L,1960L,1961L,1962L,1963L,
                                          1964L,1965L,1966L,1967L,1968L,1969L,
                                          1970L,1971L,1972L,1973L,1974L,
                                          1975L,1976L,1977L,1978L,1979L,1980L,
                                          1981L,1982L,1983L,1984L,1985L,
                                          1986L,1987L,1988L,1989L,1990L,
                                          1991L,1992L,1993L,1994L,1995L,1996L,
                                          1997L,1998L,1999L,2000L,2001L,
                                          2002L,2003L,2004L,2005L,2006L,2007L,
                                          2008L,2009L,2010L,2011L,2012L,
                                          2013L,2014L,2015L,2016L,2017L,2018L,
                                          2019L,2020L,2021L,2022L,2023L, 2024L, 2025L),
         cases = c(107473,164191,165418,152003,
                                          202210,181411,161799,197371,
                                          166914,172559,215343,179135,265269,
                                          180518,147237,214652,227319,103188,
                                          183866,222202,191383,191890,109873,
                                          133792,109860,156517,74715,69479,
                                          120718,68687,45030,37129,60886,
                                          62786,31732,28295,32148,40005,
                                          14809,11468,17749,17135,13005,6799,
                                          7717,9718,4810,3285,4249,3036,
                                          3287,1759,2402,1738,1010,2177,2063,
                                          1623,1730,1248,1895,2463,2276,
                                          3589,4195,2823,3450,4157,4570,
                                          2719,4083,6586,4617,5137,7796,6564,
                                          7405,7298,7867,7580,9771,11647,
                                          25827,25616,15632,10454,13278,
                                          16858,27550,18719,48277,28639,32971,
                                          20762,17972,18975,15609,18617,
                                          6124,2116,3044,7063, 22538, 21996 )
       )
```

I want a plot of year vs. cases
```{r}
library(ggplot2)

ggplot(cdc) +
  aes(year, cases) +
  geom_point() +
  geom_line()
```


> Q. Add annotation lines for the major milestones of wP vaccination roll-out (1946) and the switch to the aP vaccine  (1996)

```{r}
library(ggplot2)

ggplot(cdc) +
  aes(year, cases) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept = 1946, col = "blue", lty = 2)+
  geom_vline(xintercept = 1996, col = "red", lty = 2) + 
  geom_vline(xintercept = 2020, col = "grey", lty = 2)
```
It is clear from the CDC data that pertussis cases are once again increasing. For example, we can see that in 2012 the CDC reported 48,277 cases of pertussis in the United States. This is the largest number of cases reported since 1955, when 62,786 cases were reported. The pertussis field has several hypotheses for the resurgence of pertussis including (in no particular order): 1) more sensitive PCR-based testing, 2) vaccination hesitancy 3) bacterial evolution (escape from vaccine immunity), 4) waning of immunity in adolescents originally primed as infants with the newer aP vaccine as compared to the older wP vaccine.

## Exploring CMI-PB data

The CMI-PB project < https://www.cmi-pb.org > mission is to provide the scientific ommunity with a comprehensive, high-quality and freely accessible resrouce of Pertussis booster vaccination

Basically, make available a large data set on the immune response to Pertussis. They use a "booster" vaccination as a proxy for Pertussis vaccination

They make their data available as JSON format API. We can read this into R with the `read_json()` function from the **jsonlite** package:

```{r}
library(jsonlite)
subject <- read_json("https://www.cmi-pb.org/api/v5_1/subject", simplifyVector = TRUE)
head(subject)
```

> Q. How many aP and wP individuals are in this `subject` table?

```{r}
table(subject$infancy_vac)
```

How many male/female are there?
```{r}
table(subject$biological_sex)
```

> Q. What is the breakdown of `biological_gender` and `race`?

> Q. Is this representative of the US population?

```{r}
table(subject$race, subject$biological_sex)
```

We can read more tables from the CMI-PB database

```{r}
specimen <- read_json("https://www.cmi-pb.org/api/v5_1/specimen", 
                      simplifyVector = TRUE)
ab_titter <- read_json("https://www.cmi-pb.org/api/v5_1/plasma_ab_titer", 
                       simplifyVector = TRUE)
```

```{r}
head(specimen)
```

```{r}
head(ab_titter)
```
To make sense of all this data, we need to "join" (a.k.a. "merge" or "link") all these tables together. Only then will you now that a given Ab measurement (from the `ab_titter` table) was collected on a certain date (from the `specimen` table) from a certain wP or aP subject (from the `subject` table).

We can use **dplyr** and the `*_join()` family of functions to do this

```{r}
library(dplyr)

meta <- inner_join(subject, specimen)
head(meta)
```

Let's do one more `inner_join()` to join the `ab_titter` with all our meta data
```{r}
abdata <- inner_join(ab_titter, meta)
head(abdata)
```

> Q. How many different Ab "isotypes" values are in this dataset?

```{r}
table(abdata$isotype)
```


> Q. How many different "antigen" values are measured

```{r}
table(abdata$antigen)
```

Let's focus on IgG isotype

```{r}
igg <- abdata |> 
  filter(isotype =="IgG")
head(igg)
```
Make a plot of "MFI_normalized" values for all `antigen` values.

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen) +
  geom_boxplot()
```

The antigen "PT", "FIM2/3" and "FHA" appear to have the widest range of values.

> Q. Is there a difference for these responses between aP and wP individuals?

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen, col=infancy_vac) +
  geom_boxplot() 
```

```{r}
library(ggplot2)

ggplot(igg) +
  aes(MFI_normalised, antigen) +
  geom_boxplot() +
  facet_wrap(~infancy_vac)
```

Is there a difference with time (i.e. before booster shoot vs after booster shot)

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen, col=infancy_vac) +
  geom_boxplot() +
  facet_wrap(~visit)
```

Lets finish this section by looking at the 2021 dataset IgG PT antigen levels time-course:
```{r}
## Filter to 2021 dataset, IgG and PT only
ab.PT.21 <- abdata |> filter(dataset == "2021_dataset",
                              isotype == "IgG",  
                             antigen == "PT")

  ggplot(ab.PT.21) +
    aes(x=planned_day_relative_to_boost,
        y=MFI_normalised,
        col=infancy_vac,
        group=subject_id) +
    geom_point() +
    geom_line() +
    geom_smooth(aes(group = infancy_vac), 
                alpha = 0.2) +
    geom_vline(xintercept=0, linetype="dashed") +
    geom_vline(xintercept=14, linetype="dashed") +
  labs(title="2021 dataset IgG PT",
       subtitle = "Dashed lines indicate day 0 (pre-boost) and 14 (apparent peak levels)")
```

```{r}
trajectory_data <- ab.PT.21 %>%
  group_by(infancy_vac, planned_day_relative_to_boost) %>%
  summarise(mean_MFI = mean(MFI_normalised, na.rm = TRUE))
```

```{r}
ggplot() +

  # individual subjects
  geom_line(data = ab.PT.21,
            aes(x = planned_day_relative_to_boost,
                y = MFI_normalised,
                group = subject_id,
                color = infancy_vac),
            alpha = 0.3) +

  # mean trajectory
  geom_line(data = trajectory_data,
            aes(x = planned_day_relative_to_boost,
                y = mean_MFI,
                color = infancy_vac,
                group = infancy_vac),
            size = 1.5) +

  geom_point(data = trajectory_data,
             aes(x = planned_day_relative_to_boost,
                 y = mean_MFI,
                 color = infancy_vac),
             size = 3) +

  labs(x = "Days Relative to Boost",
       y = "Normalized MFI",
       color = "Infancy Vaccine") +

  theme_minimal()
```















































