---
title: "Class 7: Machine Learning I"
author: "Tonhi Dinh (A18508020)"
format: html
---

## Quarto

Today we will begin our exploration of important machine learning methods with a focus on **clustering** and **dimensionality reduction**.


To start testing these methods, let's make up some sample data to cluster where we know what the answer should be.

```{r}
hist(rnorm(3000, mean=10))
```
> Q. Can you generate 30 numbers centered at +3 and 30 numbers at -3 taken at random from a normal distribution?

```{r}
tmp <- c(rnorm(30, mean=3),
         rnorm(30, mean=-3))
x <-cbind(x=tmp, y=rev(tmp))
plot(x)
```


## K-means clustering

The main function in "base R"for K-means clustering is called `kmeans()`, let's try it out:

```{r}
k <- kmeans(x, center = 2)
k
```
> Q. What component of your kmeans result object has the cluster centers?

```{r}
k$centers
```

> Q. What component of yoru kmeans result object has the cluster size (i.e. how many points are in each cluster)?

```{r}
k$size
```

> Q. What component of your kmeans result object has the cluster membership vector (i.e. the main clustering result: which points are in which cluster)?

```{r}
k$cluster
```

> Q. Plot the results of clustering (i.e. our data colored by the clustering results) along with the cluster results?

```{r}
plot(x,col=k$cluster)
points(k$centers, col="blue", pch=15,cex=2)
```

> Q. Can you run `kmeans()` again and cluster `x` into 4 clusters and plot the results just like we did above with coloring by cluster and the cluster centers shown in blue?

```{r}
k4 <- kmeans(x, centers=4)
plot(x, col=k4$cluster)
points(k4
       $centers, col="orange", pch = 15, cex = 2)
```

> **Key points:** Kmeans will alaways reutrn the clustering that we ask for (this is the "K" or "centers" in K-means)

```{r}
k$tot.withinss
```

## Hierarchical clustering

The main function for Hierarchical clustering in base R is called `hclust()`. One of the main differences with respect to the `kmeans()` function is that you cannot just pass you input data directly from `hclust()` - it needs a "distance matrix" as input. We can get this from lots of places including the `dist()` function.

```{r}
d <- dist(x)
hc <- hclust(d)
plot(hc)
```

We can cut the dendogram or "tree" at a given height to yield our "clusters". For this we use the function `cutree()`
```{r}
plot(hc)
abline(h=4.2, col="steelblue4")
grps <- cutree(hc, h=4.2)
```

```{r}
grps
```

> Q. Plot our data `x` colored by the clustering result from `hclust()`and `cutree()`?

```{r}
grps <- cutree(hc, h=4.2)
plot(x, col=grps)
```

## Principle Component Analysis (PCA)

PCA is a popular dimensionality reduction technique that is widely used in Bioinformatics. 

### PCA for UK food data

Start the lab sheet...
```{r}
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url)
```

It looks like the row names are not set properly. We can fix this:
```{r}
rownames(x) <- x[,1]
x <- x[,-1]
```
A better way to do this is fix the row names assignment at import time:
```{r}
x <- read.csv(url, row.name = 1)
```

> Q1. How many rows and columns are in your new data frame named x? What R functions could you use to answer this questions?

```{r}
dim(x)
```

> Q2. Which approach to solving the ‘row-names problem’ mentioned above do you prefer and why? Is one approach more robust than another under certain circumstances?

```{r}
x <- x[,-1] # I prefer this one because it's simpler and involved less typing
```

>Q3: Changing what optional argument in the above barplot() function results in the following plot?

```{r}
barplot(as.matrix(x), beside=T, col=rainbow(nrow(x)))
barplot(as.matrix(x), beside=T, col=rainbow(nrow(x))) # changing beside=T to beside=F
```
> Q5: We can use the pairs() function to generate all pairwise plots for our countries. Can you make sense of the following code and resulting figure? What does it mean if a given point lies on the diagonal for a given plot?

```{r}
pairs(x, col=rainbow(nrow(x)), pch=16) # If the countries have the same consumption level, it would have the same point. The points also indicate how much each country consumes each type of food relative for each other.
```
```{r}
library(pheatmap)
pheatmap( as.matrix(x) )
```
Of all these plot really only the `pairs()` plot was useful. This however took a bit of work to interpret and will not scale when I am looking at much  bigger datasets.

## PCD the rescue

The main function in "base R" for PCA is called `prcomp()`.
```{r}
pca <- prcomp(t(x))
summary(pca)
```
> Q. How much variance is captured in the first PC?

67.4%

> Q. How many PCs do I need to capture at least 90% of the titak variance in the data set?

Two PCs captured 96.5%

> Q. Plot our main PCA results. FOlks can call this different things depending on their field of study e.g. "PC plot", "orientation plot", "score plot", "PC1 vs PC2 plot".

To generate out PCA score plot we want the pca$x component of the result object
```{r}
pca$x
```

```{r}
my_col <- c("orange", "red", "blue", "darkgreen")
plot(pca$x[,1], pca$x[,2], col=my_col)
```

```{r}
library(ggplot2)

ggplot(pca$x) + 
  aes(PC1, PC2) + 
  geom_point(col=my_col)
```



























